---
output: github_document
---


![R-CMD-check](https://github.com/mcconvil/pdxTrees/workflows/R-CMD-check/badge.svg)
![Render README](https://github.com/mcconvil/pdxTrees/workflows/Render%20README/badge.svg)




# pdxTrees

A Data Package composed of information for all the inventoried trees in Portland, OR.
There are three datasets in the package, `pdxTrees_parks`, which contains data on 25,534 trees,`ohsuTrees_parks`, which contains data on 875 trees in the parks surrounding the Oregon Health and Science University, and `pdxTrees_streets`, which contains data on 218,602 trees located on Portland's streets.
A street tree is loosely defined as a tree generally in the public right-of-way, usually between the sidewalk and the street. Street trees are categorized by one of the 96 Portland neighborhoods and the park trees are categorized by the public parks in which they grow. 



These data were collected by the Portland Parks and Recreation's Urban Forestry Tree Inventory Project.
The Tree Inventory Project has gathered data on Portland trees since 2010, collecting this data in the summer months with a team of over 1,300 volunteers and Forestry employees.
The streets trees were inventoried from 2010 to 2016, and the park trees that currently make up `pdxTrees_parks` and `ohsuTrees_parks`, were inventoried from 2017 to 2019.
More information on the data can be found [here](https://www.portlandoregon.gov/parks/article/501565).

```{r, echo= FALSE, warning = FALSE, message=FALSE}

devtools::install_github("mcconvil/pdxTrees")

library(pdxTrees)
library(tidyverse)

```

```{r, warning = FALSE}
# a histogram of the inventory date 
pdxTrees_parks %>%   
  count(Inventory_Date) %>%  
  # setting the aesthetics
  ggplot(aes(x = Inventory_Date)) +   
  # specifying a histogram and picking color! 
  geom_histogram(bins = 50,               
                 fill = "darkgreen", 
                 color = "black") + 
  labs( x = "Inventory Date", 
        y = "Count", 
        title= " When was pdxTrees_parks Inventoried?") + 
  # adding a theme 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))

```





`pdxTrees` is used in multiple Reed College statistics courses as a tool to help students learn and become more comfortable with the functionality of RStudio.
In Intro to Probability and Statistics, `pdxTrees` is used to learn the components of graphics, wrangle data with `dplyr`, and create graphs with `ggplot2`.
In the Data Science course, the package is used to teach the best practices of function writing, construct interactive maps with `leaflet`, and showcase the usefulness of `lubridate` along with many other R packages. 




# Install the package

```{r, message=FALSE, warning = FALSE}
# Do the following once
# install.packages("devtools")

# Then install the package
devtools::install_github("mcconvil/pdxTrees")
```

## Update 
`pdxTrees` was update on June 25th  to contain three data sets `pdxTrees_parks`, `pdxTrees_streets`, and `ohsuTrees_parks`.
New park data from 2019 was added in addition to all of the street tree data. Enjoy! 


