---
output: github_document
---


![R-CMD-check](https://github.com/mcconvil/pdxTrees/workflows/R-CMD-check/badge.svg)
![Render README](https://github.com/mcconvil/pdxTrees/workflows/Render%20README/badge.svg)




# pdxTrees

A Data Package composed of information on inventoried trees in Portland, OR.
There are two datasets in the package: 

* `pdxTrees_parks`, which contains data on 25,534 trees from 174 Portland parks.

* `pdxTrees_streets`, which contains data on 218,602 trees located on Portland's streets. A street tree is loosely defined as a tree generally in the public right-of-way, usually between the sidewalk and the street. 

The street trees are categorized by one of the 96 Portland neighborhoods and the park trees are categorized by the public parks in which they grow. 


These data were collected by the [Portland Parks and Recreation's Urban Forestry Tree Inventory Project](https://www.portlandoregon.gov/parks/53181). The Tree Inventory Project has gathered data on Portland trees since 2010, collecting this data in the summer months with a team of over 1,300 volunteers and city employees. The streets trees were inventoried from 2010 to 2016, and the park trees that currently make up `pdxTrees_parks`, were inventoried from 2017 to 2019. More information on the data can be found [here](https://www.portlandoregon.gov/parks/article/501565).

```{r, echo= FALSE, warning = FALSE, message=FALSE}
devtools::install_github("mcconvil/pdxTrees")
```

```{r, warning = FALSE, message = FALSE}
library(pdxTrees)
library(tidyverse)

# Grabing the data 

pdxTrees_parks <- get_pdxTrees_parks()

# Histogram of the inventory date 
pdxTrees_parks %>%   
  count(Inventory_Date) %>%  
  # Setting the aesthetics
  ggplot(aes(x = Inventory_Date)) +   
  # Specifying a histogram and picking color! 
  geom_histogram(bins = 50,               
                 fill = "darkgreen", 
                 color = "black") + 
  labs( x = "Inventory Date", 
        y = "Count", 
        title= " When was pdxTrees_parks Inventoried?") + 
  # Adjusting the theme 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))

```



## To install the package

```{r, message=FALSE, warning = FALSE}
# Do the following once
# install.packages("devtools")

# Then install the package
devtools::install_github("mcconvil/pdxTrees")
```


## Update/Getting the data 

`pdxTrees` was updated in July of 2020 to contain two data loading functions, `get_pdxTrees_parks()` and `get_pdxTrees_streets` to pull the park and street tree data, respectivly, from the github repository.

```{r, warning=FALSE, message=FALSE}
# To get data on all parks 

  pdxTrees_parks <- get_pdxTrees_parks()

# To get data on one park 

  berkeley_park <- get_pdxTrees_parks(park = "Berkeley Park")

# Multiple parks 

  parks <- get_pdxTrees_parks(park = c("Berkeley Park", "East Delta Park"))


# The streets function works the same way but with neighborhoods! 

  pdxTrees_streets <- get_pdxTrees_streets()

# One neighborhood 
  
  concordia <- get_pdxTrees_streets(neighborhood = "Concordia")
  
# Mutliple neighborhoods! 
  
  neighborhoods <- get_pdxTrees_streets(neighborhood = c("Concordia", "Eastmoreland", "Sunnyside"))

```
## Teaching with `pdxTrees`

`pdxTrees` is used in multiple [Reed College statistics courses](https://www.reed.edu/math/courses.html) as a tool to help students develop their data analysis skills in [RStudio](https://rstudio.com/). In Introduction to Probability and Statistics, students wrangle `pdxTrees` data with `dplyr`, and create graphs of `pdxTrees` with `ggplot2`. In the [Data Science course](https://github.com/Reed-Statistics/math241s20), the package is used to teach the best practices of function writing, to construction interactive maps with `leaflet`, and to showcase the usefulness of `lubridate` along with many other R packages. 



