---
title: "Data Wrangling with pdxTrees"
output: rmarkdown::html_vignette 


vignette: >
  %\VignetteIndexEntry{Data Wrangling with pdxTrees}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  
---

```{r, include = FALSE}
knitr::opts_chunk$set( echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>",  fig.align = "center",
                       fig.retina = 2)
```


`pdxTrees` is a data package composed of information on inventoried trees in Portland, OR. There are two datasets that can be accessed with this package:

* `get_pdxTrees_parks(version = "2019")` pulls in data on up to 25,534 trees from 174 Portland parks.
* `get_pdxTrees_parks(version = "2026")` pulls in data on up to 25,734 trees from 316 Portland parks.

* `get_pdxTrees_streets(version = "2019")` pulls in data on up to 218,602 trees located on Portland's streets. A street tree is loosely defined as a tree generally in the public right-of-way, usually between the sidewalk and the street. 
*`get_pdxTrees_streets(version = "2026)` pulls in data on up to 252,180 trees located on Portland's streets. A street tree is loosely defined as a tree generally in the public right-of-way, usually between the sidewalk and the street. 

The street trees are categorized by one of the 95 Portland neighborhoods and the park trees are categorized by the public parks in which they grow. 

Here are some examples of the different ways `pdxTrees` can be used in an educational setting to learn data visualization!

 
```{r, include = TRUE, messgae = FALSE}

# First make sure you have the package downloaded! 

devtools::install_github("mcconvil/pdxTrees")

# Loading the required libraries

library(pdxTrees) 
library(ggplot2)
library(dplyr)
library(forcats)
library(devtools)

```


## Creating a new variable with mutate() and case_when()

The following code creates an new variable with `mutate`, `case_when`, and the `pdxTrees_parks` data. It:
* Removes three variables from the new subset data set using `select`
* Creates a new variable that groups the carbon storage values int NA, Low, Moderate, and High
* Shows you the first few rows of the new subset data set
* Counts how many trees are in each category of `carbon_group`

```{r}

# Pulling the data for Parks from the 2019 data set

pdxTrees_Parks <- get_pdxTrees_parks()

#create a subset data frame
Carbon_Parks <- pdxTrees_Parks %>%
	#get rid of extra variables
	select(-c("Collected_By", "Nuisance", "Functional_Type")) %>%
	#add a new variable that groups carbon storage value
	mutate(Carbon_Group = case_when(
		is.na(Carbon_Storage_value) ~ NA,
		Carbon_Storage_value <= 250 ~ "Low",
		between(Carbon_Storage_value, 250, 750) ~ "Moderate",
		Carbon_Storage_value > 750 ~ "High")) 


#view small piece of new data frame
glimpse(Carbon_Parks)


#show how many observations in each carbon group
#note there are 49 NAs to begin with
count(Carbon_Parks, Carbon_Group)

```


## Finding Summary Statistics and grouping

The following code finds summary statistics for each category of `Condition` before arranging it in ascending order with `group_by`, `summarize`, `arrange`, and the `pdxTrees_parks` data. 

```{r}

#create a subset data frame
Park_Stats <- Carbon_Parks %>%
  # group data by each category of condition
	group_by(Condition) %>%
  # find the mean DBH and tree height in each category and count how many trees per category
	summarize(mean_DBH = mean(DBH, na.rm = TRUE),
		mean_tree_height = mean(Tree_Height, na.rm = TRUE),
		trees = n()) %>%
  #arrange the data in ascending order by trees
	arrange(trees)


# show table output
Park_Stats

```

## Changing the class of a variable and ordering the levels

The following code identifies the class and level of a variable and changes it with `class`, `mutate`, `factor`, `levels`, and the `pdxTrees_parks` data.

```{r}

#show variable class 
class(Carbon_Parks$Condition)
#show variable order
levels(Carbon_Parks$Condition)


#change to ordered and set the order
Carbon_Parks <- Carbon_Parks %>%
  mutate(Condition = factor(Condition,
      levels = c("Dead", "Poor", "Fair", "Good")))


#show new variable class 
class(Carbon_Parks$Condition)
#show new variable order
levels(Carbon_Parks$Condition)

```

## Using the or logical operator to select above average trees

The following code filters the data based on selected variable values with `filter` and the `pdxTrees_parks` data.

```{r}

#create a subset data frame
Carbon_Parks <- Carbon_Parks %>%
	#show rows where condition is set to good and tree height is greater than or equal to its average or DBH is greater than or equal to its average
filter(Condition == "Good", (Tree_Height >= 50 | DBH >= 17.5))


# show table output
Carbon_Parks

```

## Simple random sampling

The following code creates an simple random sample with `slice_sample`, and the `pdxTrees_parks` data.

```{r}

#create a subset data frame
Park_Sample <- Carbon_Parks %>%
	#simple random sample
	slice_sample(n = 100)


#display first few rows
head(Park_Sample)

```


## Summarize 

The following code summarizes statistic your sampled dataset with `select`, `summary`, `print`, and the `pdxTrees_parks` data.

```{r}

#create a name to store data under
Summary_Sample <- Park_Sample %>% 
  #choose variables to summarize
  select(DBH, Tree_Height) %>% 
  #get summary statistics on chosen variables
  summary()


#show output and observe changes to medians
Summary_Sample

```
## Convert `DBH` and `Tree_Height` to the Metric System 

The following code converts `DBH` and `Tree_Height` to the metric system with `mutate`, `glimpse`, and the `pdxTrees_parks` data.
```{r}

pdxTrees_Parks_m <- pdxTrees_Parks %>% 
  # multiply each variable by the conversion and change `DBH` to cm and `Tree_Height` to m
  mutate(DBH_cm = 2.54 * DBH,
         Tree_Height_m = 0.3048 * Tree_Height)


#show table
glimpse(pdxTrees_Parks_m)


# Pulling the data for Streets from the 2019 data set
pdxTrees_Streets <- get_pdxTrees_streets()


pdxTrees_Streets_m <- pdxTrees_Streets %>% 
  # multiply each variable by the conversion and change `DBH` to cm and `Tree_Height` to m
  mutate(DBH_cm = 2.54 * DBH)


#show table
glimpse(pdxTrees_Streets_m)

```

## Are Nuisance trees commonly the inedible or edible trees?

The following code test if Nuisance trees commonly the inedible or edible trees with `filter`, `count`, and the `pdxTrees_parks` data.
```{r}

# filter out NAs
pdxParks_na <- pdxTrees_Parks %>% 
  filter(Edible == c("No", "Yes", "Yes - fruit", "Yes - nuts"))

# filter nuisance to yes
nuisance_parks <- pdxParks_na %>% 
  filter(Nuisance == "Yes") 
 

# use count to compare the two
count(nuisance_parks, Edible)
count(pdxParks_na, Edible)

```

## What Neighborhoods have the most high voltage wires near trees

The following code shows how many high voltage wires are near trees in each neighborhood with `filter`, `group_by`, `count`, and the `pdxTrees_parks` data.
```{r}

# grouping to the variables we want to look at and making sure were only looking at high voltage
pdxTrees_Streets %>% 
  group_by(Neighborhood, Wires) %>% 
  filter(Wires == "High voltage") %>% 
  #count high volt wires in each neighborhood
  count(., Wires) %>% 
  #arrange in descending order
  arrange(desc(n))

#maybe add in 2026 comparison

```

## Do big trees tend to grow on larger sites

The following code compares DBH with site size with `filter`, `group_by`, `count`, and the `pdxTrees_parks` data.
```{r}

# grouping to the variables we want to look at and making sure were only looking at high voltage
dbh_size <- pdxTrees_Streets %>% 
  group_by(Site_Size) %>% 
  summarise(avg_dbh = mean(DBH, na.rm = TRUE),
         trees = n())


dbh_size


#add comparison here to 2026 data

```

