---
title: "Data Visulization with pdxTrees"
output: rmarkdown::html_vignette 


vignette: >
  %\VignetteIndexEntry{Data Visulization with pdxTrees}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}

---

```{r, include = FALSE}
knitr::opts_chunk$set( echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>",  fig.align = "center",
                       fig.retina = 2)
```


`pdxTrees` is a data package composed of information on inventoried trees in Portland, OR. There are two datasets that can be accessed with this package:

* `get_pdxTrees_parks(version = "2019")` pulls in data on up to 25,534 trees from 174 Portland parks.
* `get_pdxTrees_parks(version = "2026")` pulls in data on up to 25,734 trees from 316 Portland parks.

* `get_pdxTrees_streets(version = "2019")` pulls in data on up to 218,602 trees located on Portland's streets. A street tree is loosely defined as a tree generally in the public right-of-way, usually between the sidewalk and the street. 
*`get_pdxTrees_streets(version = "2026)` pulls in data on up to 252,180 trees located on Portland's streets. A street tree is loosely defined as a tree generally in the public right-of-way, usually between the sidewalk and the street. 

The street trees are categorized by one of the 95 Portland neighborhoods and the park trees are categorized by the public parks in which they grow. 

Here are some examples of the different ways `pdxTrees` can be used in an educational setting to learn data visualization!

 
```{r, include = TRUE, messgae = FALSE}

# First make sure you have the package downloaded! 

devtools::install_github("mcconvil/pdxTrees")

# Loading the required libraries

library(pdxTrees)
library(dplyr)
library(forcats)
library(devtools)
library(tidyverse)
library(ggalluvial)
library(leaflet)
library(leaflet.extras)
```
## Load data into environment using a package function

```{r}
PDX_parks <- get_pdxTrees_parks()
PDX_streets <- get_pdxTrees_streets()
```

# Create a histogram to explore Portland park tree heights

The following:
* Uses the data that was loaded using a package function
* Makes a colored histogram to display tree heights
* Makes 30 bins for data

```{r}
#plotting

ggplot(data = PDX_parks,
 mapping = aes(x = Tree_Height)) +
 geom_histogram(color = "white",
 fill = "forestgreen",
 bins = 30) +
  labs(title = "Distribution of Tree Heights in Portland Parks")

```
# Create a histogram to explore Portland street tree DBH

The following:

* Uses the data that was loaded using a package function
* Makes a colored histogram to display tree DBH on Portland Streets
* Makes 45 bins for data

```{r}
#plotting

ggplot(data = PDX_streets,
 mapping = aes(x = DBH)) +
 geom_histogram(color = "white",
 fill = "green3",
 bins = 45) +
  labs(title = "Distribution of Tree DBH in Portland Streets")
```

# Is DBH associated with condition in street trees?

The following:
* Factors in order to list condition in a desired order
* Adds labels
* Utilizes geom_violin


```{r}
#plotting

ggplot(
  data = PDX_streets,
  mapping = aes(
    x = factor(Condition, levels = c("Dead", "Poor", "Fair", "Good")),
    y = DBH,
    fill = factor(Condition, levels = c("Dead", "Poor", "Fair", "Good"))
  )
) +
  geom_violin() +
  scale_fill_brewer() +
    labs(x = "Condition", fill = "Condition", title = "DBH with each Tree Condition Level")
```

# Explor DBH and carbon storage value in Portland parks

The following:
* Creates a data frame of cleaned data containing Carbon_Storage_value and DBH, also removing NA values
* Uses geom_jitter to plot DBH against Carbon Storage Values
* Uses geom_smooth to identify trends


```{r}
#cleaning

Carbonstorage_DBH <- PDX_parks %>%
  select(Carbon_Storage_value, DBH) %>%
  drop_na()

#plotting

ggplot(data = Carbonstorage_DBH, mapping = aes(x = Carbon_Storage_value, y = DBH)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth() +
  labs(title = "DBH and Carbon Storage Value")
```
# Does site width correlate with DBH for Hophornbeam on Portland streets?

The following:

* Creates a data frame using select() and filter() to subset desired data
* Generates a line graph to display that DBH and site width don't increase together
```{r}
#cleaning
street_hophorn <- PDX_streets %>%
  select(Common_Name, DBH, Site_Width) %>%
  filter(Common_Name %in% "Hophornbeam") %>%
  drop_na()

#plotting
ggplot( data = street_hophorn, mapping = aes(x = Site_Width, y = DBH)) +
  geom_line(color = "darkseagreen") +
  labs(title = "Hophornbeam DBH and Site Width Relationship" )
```


# Carbon storage of Portland park trees considering various sizes

The following: 

* Cleans creates a data frame containing the data which includes variables of interest, including Carbon_Storage_value and Mature_Size
* Plots the data using geom_density


```{r}
#cleaning

parks_density <- PDX_parks %>%
  drop_na(Carbon_Storage_value, Mature_Size)

#plotting

ggplot(data =  parks_density, mapping = aes(x = Carbon_Storage_value, fill = Mature_Size)) +
  geom_density() +
  labs( title = "Carbon Storage of Various Tree Sizes")
```

# Did staff or volunteers collect more data in Parks?

This graph observes data on who collected data in Portland parks to see which 
group collected more data.

The following:

* Removes NA values in the data that counts who collected the Park tree data
* Uses a bar graph with different colors to compare volunteer and staff 
collection rates

```{r}
#cleaning

no_na_parks <- PDX_parks %>%
  drop_na(Collected_By)

#plotting

ggplot( data = no_na_parks, mapping = aes(x = Collected_By, fill = Collected_By)) +
  geom_bar(color = "black") +
  theme_bw() +
  scale_fill_manual(values = c("darkgoldenrod4","darkcyan")) +
  labs(title = "Staff vs. Volunteer Collection in Portland Parks")
```


# Did staff or volunteers collect more data on Streets?

The following:

* Removes NA values in the data that counts who collected the street tree data
* Uses a bar graph with different colors to compare volunteer and staff 
collection rates

```{r}
#cleaning

no_na_streets <- PDX_streets %>%
  drop_na(Collected_By)

#plotting

ggplot( data = no_na_streets, mapping = aes(x = Collected_By, fill = Collected_By)) +
  geom_bar(color = "black") +
  theme_bw() +
  scale_fill_manual(values = c("darkkhaki","azure1")) +
   labs(title = "Staff vs. Volunteer Collection on Portland Streets")
```
# How many of these species of interest are there in Portland parks?

The following:

* Makes a data frame to select and filter to data of interest
* Makes counts for tree species of interest
* Plots using geom_col() with a color palette
```{r}
tree_species_count <- PDX_parks %>%
   select("Common_Name") %>%
  filter(Common_Name %in% c("Douglas-Fir",
                             "Giant Sequoia",
                             "Sweetgum")) 
 

counts_spe <- count(tree_species_count, Common_Name)

ggplot(data = counts_spe, mapping = aes(x = Common_Name, y=n, fill = Common_Name)) +
  geom_col() +
  scale_fill_brewer(palette = "PuBuGn") +
  labs(title = "Douglas Fir, Giant Sequioa, and Sweetgum Presence in Portland Parks")
```


# Do trees in better conditions sequester more carbon?

The following:

* Creates a data frame containing condition and wires from Portland Streets
* Uses count() to make a data frame that counts condition and wires
* Plots the information using geom_alluvium


```{r}
#cleaning

condition_wires <- get_pdxTrees_streets() %>%
    select("Wires", "Condition")

#counting

counts_wire_con <- count(condition_wires, Condition, Wires)

#plotting

ggplot(data = counts_wire_con,
       mapping = aes( axis1 = Condition, axis2 = Wires, y = n)) +
  geom_alluvium(mapping = aes(fill = Condition)) +
  geom_stratum() +
  geom_text(stat = "stratum", 
            mapping = aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Condition", "Wires")) +
  labs(title = "Wires Effect on Tree Condition")
```


# Create a bar graph comparing native and nonnative trees in Portland parks

This bar graph compares native tree species count to nonnative tree species count in Portland parks.

The following:
* Removes NA values in the data that counts native and nonnative trees
* Makes a bar graph comparing native and nonnative tree counts
* Uses a color palette to differentiate

```{r}
#cleaning

native_parktrees <- PDX_parks %>%
  drop_na(Native) 

#plotting

ggplot(data = native_parktrees,
 mapping = aes(x = Native, fill = Native)) +
 geom_bar(color = "black") +
  theme_bw() +
  labs( title = "Native vs Nonnative Tree Count in Portland Parks",
        y = "Number of Trees") +
  scale_fill_manual(values = c("red3","lightgreen"))
```


# Using geom_alluvium to see the condition of Ash/Mountain Ash trees in 2026 Portland streets

The following:
* Filters to two types of ash trees from the streets dataset from 2026
* Selects species and condition
* Creates a data frame with count to analyze common observations (provides "n" column)
* Makes a geom_alluvium to analyze the condition of ash trees on Portland streets
* Uses geom_stratum to draw rectangular blocks on each axis
* Renames the axes
```{r}

#cleaning
ash_trees_condition <- get_pdxTrees_streets(version = "2019") %>%
  filter(Common_Name %in% c("Mountain-Ash", 
                            "Ash")) %>%
  select("Common_Name", "Condition")

#counting

counts <- count(ash_trees_condition, Common_Name, Condition)

#plotting

ggplot(data = counts, mapping = aes( axis1 = Common_Name, axis2 = Condition, y = n)) +
  geom_alluvium(mapping = aes(fill = Common_Name)) +
  geom_stratum() +
  geom_text(stat = "stratum", mapping = aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Species", "Condition")) +
  labs(title = "Condition of Ash Trees on Portland Streets in 2026")
```
# Comparing street tree diameters based on family

The following:

* Creates a data frame that selects and filters data of interest
* Uses a violin plot to compare DBH based on Tree Families
```{r}
#cleaning

street_family_eval <- PDX_streets %>%
  select(Family, DBH) %>%
   filter(Family %in% c("Fagaceae",
                             "Sapindaceae",
                             "Rosaceae"))

#plotting

ggplot( data = street_family_eval,
        mapping = aes(x = Family,
                      y = DBH,
                      fill = Family)) +
  geom_violin() +
  labs(title = "Tree Family Differences in DBH")
```


# Comparing park tree diameters based on species

The following:
* Creates a data frame that subsets down to species of interest and DBH
* Makes a boxplot comparing the DBH of different species of street trees
* Uses a color palette


```{r}
#cleaning

Parks_species_eval <- PDX_parks %>%
  select(Common_Name, DBH, Park) %>%
   filter(Common_Name %in% c("Douglas-Fir",
                             "Giant Sequoia",
                             "Sweetgum"), 
    Park %in% c(
      "Kenilworth Park",
      "Berkeley Park",
      "Westmoreland Park",
      "Woodstock Park"))

#plotting

ggplot( data = Parks_species_eval,
        mapping = aes(x = Common_Name,
                      y = DBH,
                      fill = Common_Name)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "PiYG") +
  facet_wrap(~Park) +
  labs(title = "DBH of Various Tree Species in Parks near Reed College")
```
# Which park is best for my favorite snack?

The following: 
* Selects park and edible status for a new data frame
* Drops NA values for selected columns
* Removes "no" values from Edible and selects 6 different parks for analysis
* Plots data, facets by park

```{r}
#cleaning

edible_parks <- PDX_parks %>%
  select(Park, Edible) %>%
  drop_na() %>%
  filter(Edible != "No", 
         Park %in% c(
      "Ed Benedict Park",
      "Midland Park",
      "Columbia Childrens Arboretum",
      "Fernhill Park",
      "Frazer Park",
      "Buckman Field"))
  
#plotting

ggplot( data = edible_parks, mapping = aes(x = Edible, fill = Edible)) +
  geom_bar(color = "black") +
  scale_fill_brewer(palette = "PRGn") +
  facet_wrap(~Park) +
labs(title = "Parks with Edible Fruit and Nut Access")

```

# Does DBH and family effect the amount of carbon sequestration in Farragut Park?

The following:
* Makes a data frame for Farragut Park
* Makes a graph using geom_jitter
* Uses geom_smooth to observe trends in the points using linear model
* Colors by Family

```{r}

#sourcing

farragut_prk <- get_pdxTrees_parks(park = "Farragut Park")

#plotting

ggplot(data = farragut_prk,
       mapping = aes(x = Carbon_Sequestration_lb, y = DBH,
                     color = Family)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "How does Carbon Sequestration Vary by Tree Family?")
```


# Comparing ash trees on Portland streets from 2017 to 2026

The following:
* Selects species and tree condition categories 
* Makes every observation in Common_Name lowercase
* Uses str_detect() to check for all ash trees in Common_Name
* Repeats process for 2017 and 2026 versions
* Plots both graphs
```{r}
#cleaning
streets_2026 <- get_pdxTrees_streets() %>%
   select(Common_Name, Condition)

Streets_2026_cleaned <- streets_2026 %>% 
  filter(str_detect(str_to_lower(Common_Name), "ash"))

streets_2019 <- get_pdxTrees_streets(version = "2019") %>%
   select(Common_Name, Condition)

Streets_2019_cleaned <- streets_2019 %>% 
  filter(str_detect(str_to_lower(Common_Name), "ash"))

#plotting

ggplot(data = Streets_2019_cleaned, mapping = aes(x = Common_Name)) +
  geom_bar() +
  facet_wrap(~Condition) +
  labs(title = "Ash Tree Counts per Condition 2019")
  
ggplot(data = Streets_2026_cleaned, mapping = aes(x = Common_Name)) +
    geom_bar() +
  facet_wrap(~Condition) +
  labs(title = "Ash Tree Counts per Condition 2026")
```



## An interactive map with `leaflet`

The following code creates an interactive map with `leaflet` and the `pdxTrees_parks` data. It showcases:

* Adding popups for each tree where you can customize the displayed information.
* Changing the background map.
* Including a mini-map.
* Allows for analysis of updates pdxTrees data
* Works with the latest available data in 2026


```{r leaflet graph, fig.width= 8, fig.height=6}

# Making the leaf popup icon 
greenLeaflittle <- makeIcon(
  iconUrl = "https://leafletjs.com/examples/custom-icons/leaf-green.png",
  iconWidth = 10, iconHeight = 20,
  iconAnchorX = 10, iconAnchorY = 10,
  shadowUrl = "https://leafletjs.com/examples/custom-icons/leaf-shadow.png",
 shadowWidth = 10, shadowHeight = 15,
 shadowAnchorX = 5, shadowAnchorY = 5
)


# Pulling the data for Berkely Park 

berkeley_prk <- get_pdxTrees_parks(version = "2026", park = "Berkeley Park")


# Creating the popup label 

labels <- paste("</b>", "Common Name:",
                 berkeley_prk$Common_Name,
                 "</b></br>", "Factoid: ", 
              berkeley_prk$Species_Factoid) 


leaflet() %>%
  # Center the map
  setView(lng = -122.6239, lat = 45.4726, zoom = 17) %>%
  
  # Base maps
  addProviderTiles(providers$Esri.WorldTopoMap,
    group = "Topographic") %>%
  
  addProviderTiles(providers$Esri.WorldImagery,
    group = "Satellite") %>%
  
  # Tree markers
  addMarkers(
    ~Longitude,
    ~Latitude,
    data = berkeley_prk,
    icon = greenLeaflittle,
    popup = ~labels
  ) %>%
  
  # Layer selector
  addLayersControl(baseGroups = c("Topographic", "Satellite"),
    options = layersControlOptions(collapsed = FALSE)) %>%
  
  # Mini map
  addMiniMap()
``` 

